---
title: "ACES: Production Frontiers, Technologies, and Efficiency"
author: "Víctor J. España"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    number_sections: true
bibliography: ../inst/REFERENCES.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{ACES: Production Frontiers, Technologies, and Efficiency}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
options(digits = 4)
```

```{r library}
library(aces)
```

# Why use ACES?

Suppose that several organizations use comparable resources to produce
comparable outcomes. A conventional regression describes the output of an
average organization. For benchmarking, that is usually not the quantity of
interest: we want to know how much output appears attainable with a given input
bundle, and how far each organization is from that benchmark.

ACES estimates this upper production frontier with adaptive regression splines.
It then uses the fitted outputs to build a production technology and solves an
envelopment model to obtain efficiency scores and targets. These three objects
play different roles:

1. the **frontier model** estimates maximum attainable outputs;
2. the **technology** defines the feasible input-output set;
3. the **efficiency model** projects each unit onto that set.

Keeping these roles separate makes both the statistical model and the
management interpretation clearer.

```{r workflow, eval=FALSE}
fit        <- aces(data, x = input_columns, y = output_columns)
technology <- get_technology(fit)
scores     <- get_scores(data, input_columns, output_columns, object = fit)
targets    <- get_targets(data, input_columns, output_columns, object = fit)
```

The package provides three estimators. Standard ACES is the natural starting
point when the fitted spline structure matters. Quick ACES searches a smaller
set of variables and knots. RF-ACES averages randomized ACES learners and is
useful when stability, prediction, or interval diagnostics are priorities.

| Estimator | Best suited to | Main trade-off |
|:--|:--|:--|
| ACES | A detailed, interpretable frontier specification | The full candidate search can be expensive |
| Quick ACES | Fast exploration or a high-dimensional search | Screening may select a different model |
| RF-ACES | Stable ensemble estimates and uncertainty diagnostics | More fitting and storage |

```{r estimator-choice, eval=FALSE}
fit_aces  <- aces(data, x, y)
fit_quick <- aces(data, x, y, quick_aces = TRUE)
fit_rf    <- rf_aces(data, x, y, learners = 100)
```

# Preparing the data

## Define comparable decision-making units

Each row is a decision-making unit (DMU): a hospital, school, branch, plant, or
other producer. Inputs are resources or commitments and outputs are desirable
results. The units should operate under a sufficiently similar mandate and
environment for a common frontier to be meaningful. Otherwise, structural
heterogeneity can easily be mistaken for inefficiency.

The package expects numeric input and output columns. Variable names are part of
the fitted model, so evaluation data must use the same names as the training
data.

```{r data-contract, eval=FALSE}
production_data <- data.frame(
  labour  = c(12, 15, 11, 19),
  capital = c(20, 18, 15, 24),
  output  = c(31, 35, 26, 43),
  row.names = paste0("DMU", 1:4)
)

fit <- aces(production_data, x = 1:2, y = 3)
```

Inputs and desirable outputs should normally be non-negative and have a
consistent direction across DMUs. Missing values should be resolved before
fitting. Percentage and logarithmic error metrics also require particular care
when an output can be zero.

```{r check-data, eval=FALSE}
x <- match(c("x1", "x2", "x3"), names(production_data))
y <- match("y", names(production_data))

stopifnot(
  all(complete.cases(production_data[, c(x, y)])),
  all(as.matrix(production_data[, c(x, y)]) >= 0)
)
```

## A reproducible example

The following generator creates three inputs, an observed output `y`, and its
latent frontier value `yT`. The latter is available because this is a simulation;
it must not be passed to the estimator as if it were observed in an empirical
application.

```{r simulated-data}
set.seed(314)
production_data <- cobb_douglas_XnY1(N = 50, nX = 3)
head(production_data)
```

A useful way to think about the single-output model is

\[
y_i=f(x_i)+v_i-u_i,
\]

where $f(x_i)$ is maximum attainable output, $u_i\geq0$ is technical
shortfall, and $v_i$ is statistical noise when a stochastic interpretation is
appropriate. In the deterministic version, $v_i=0$ and $y_i\leq f(x_i)$.
The usual production assumptions make $f$ non-decreasing and concave: more
input cannot reduce maximum output, and marginal products do not increase.

## Scaling

By default, `scale_data = TRUE` divides every model variable by its training
mean before estimation. This improves numerical conditioning without changing
the units in the results: technologies, predictions, scores, and targets are
returned on the original scale. This is mean scaling, not z-standardization.

```{r scaling, eval=FALSE}
fit_scaled <- aces(
  production_data,
  x = 1:3,
  y = 4,
  scale_data = TRUE
)

fit_unscaled <- aces(
  production_data,
  x = 1:3,
  y = 4,
  scale_data = FALSE
)
```

# Fitting an ACES frontier

## Start with the default model

For a first fit, only the data and column positions are required. Printing the
object gives a compact model summary, while `names(fit$methods)` shows the
fitted representations stored for later use.

```{r basic-fit, eval=FALSE}
fit <- aces(
  data = production_data,
  x = 1:3,
  y = 4
)

fit
names(fit$methods)
```

Internally, ACES first builds a deliberately rich spline model, then prunes it,
and finally creates smooth cubic and quintic versions. The corresponding method
names are `"aces_forward"`, `"aces"`, `"aces_cubic"`, and
`"aces_quintic"`.

```{r fitted-methods, eval=FALSE}
tech_forward <- get_technology(fit, method = "aces_forward")
tech_linear  <- get_technology(fit, method = "aces")
tech_cubic   <- get_technology(fit, method = "aces_cubic")
tech_quintic <- get_technology(fit, method = "aces_quintic")
```

Use the same method name when comparing a technology, its efficiency scores,
and its targets. Otherwise, the analysis quietly changes the benchmark.

## How the spline adapts to the data

For input $x_j$ and knot $t$, ACES works with the reflected hinge pair

\[
(x_j-t)_+=\max(x_j-t,0), \qquad (t-x_j)_+=\max(t-x_j,0).
\]

Each hinge is active on one side of the knot. By combining several hinges, the
model can change marginal products where the data support a change instead of
imposing one global functional form [@friedman1991; @espana2024]. Products of
hinges introduce interactions between inputs.

During the forward stage, every eligible input-knot candidate is assessed and
the best reflected pair is added. `max_terms` is an upper bound on the search;
`err_red` is the minimum relative improvement required to continue. Thus, a
smaller `err_red` and a larger `max_terms` allow a longer and more expensive fit.

```{r forward-controls, eval=FALSE}
fit_flexible <- aces(
  production_data,
  x = 1:3,
  y = 4,
  metric = "mse",
  max_terms = 80,
  err_red = 0.005
)
```

The fitting metric can be `"mae"`, `"mape"`, `"mse"`, `"msle"`, `"rmse"`,
`"nrmse1"`, or `"nrmse2"`. Squared-error metrics give large misses more
influence; absolute-error metrics are less sensitive to them. Percentage and log
metrics should only be used when their denominator and logarithm are meaningful.

## Pruning an intentionally rich model

The forward model is not intended to be the final answer. ACES removes eligible
basis functions one at a time and compares the resulting sequence with
generalized cross-validation,

\[
\operatorname{GCV}(M)=
\frac{\operatorname{LOF}(M)}
{\left[1-\{n_{BF}+d\,n_K\}/(Nq)\right]^2}.
\]

Here $q$ is the number of outputs, $n_{BF}$ is the basis count, $n_K$ is
the degree-weighted knot count, and $d=\texttt{kn_penalty}$. A larger knot
penalty favors a simpler final model.

```{r pruning, eval=FALSE}
fit_simple <- aces(
  production_data,
  x = 1:3,
  y = 4,
  kn_penalty = 3
)
```

The piecewise-linear model may have visible kinks. Its cubic and quintic
versions replace each local kink with a differentiable transition. Smoothing is
useful when marginal products are part of the interpretation; it does not change
the need to state which method supplied the final benchmark.

## Monotonicity and concavity

With the default `shape`, interval slopes are constrained to be non-negative and
non-increasing. The fitted spline can then be interpreted as a non-decreasing,
concave production function. These restrictions are substantive assumptions,
not cosmetic regularization.

```{r shapes, eval=FALSE}
fit_shape <- aces(
  production_data,
  x = 1:3,
  y = 4,
  shape = list(mono = TRUE, conc = TRUE)
)

fit_relaxed <- aces(
  production_data,
  x = 1:3,
  y = 4,
  shape = list(mono = FALSE, conc = FALSE)
)
```

When either restriction is relaxed, the spline is better viewed as an
intermediate predictor of maximum output. The final DEA-type envelopment still
restores free disposability and, under variable returns to scale, convexity. A
relaxed spline and a valid final technology are therefore compatible, but they
are not the same mathematical object.

## Interactions and candidate knots

`mul_BF$max_degree = 1` gives an additive frontier. Setting it to two allows
pairwise interactions, while `inter_cost` requires an interaction to improve on
the best first-degree candidate by an additional amount. A larger cost makes
interactions harder to enter. Their presence shows that the fitted frontier uses
joint input information; it is not evidence of a causal complementarity.
ACES labels interaction variables with `:`, following standard statistical
notation: for example, `capital:labour` is the product of those two inputs.

```{r interactions, eval=FALSE}
fit_interactions <- aces(
  production_data,
  x = 1:3,
  y = 4,
  mul_BF = list(
    max_degree = 2,
    inter_cost = 0.05
  )
)
```

Automatic knot selection uses `minspan` and `endspan` to keep candidates apart
and away from the boundaries. A custom `kn_grid` can be useful when the input
scale has meaningful breakpoints or when a coarser search is needed for speed.
It must contain one numeric vector for each original input.

```{r knot-controls, eval=FALSE}
custom_grid <- lapply(production_data[1:3], function(z) {
  as.numeric(quantile(z, probs = seq(0.1, 0.9, by = 0.1)))
})

fit_grid <- aces(
  production_data,
  x = 1:3,
  y = 4,
  kn_grid = custom_grid,
  minspan = 3,
  endspan = 2
)
```

# What optimization problem is being solved?

Every candidate basis needs coefficients before its fit can be compared with
the current model. These coefficients are not obtained by ordinary least
squares. ACES solves a linear program that keeps the fitted frontier above the
relevant frontier observations and imposes the selected shape restrictions.

For output $r$, let $B_E$ be the candidate basis matrix for observations on
the output-specific FDH envelope, $y_E$ their observed output, and
$w_i=1/\phi_i^{DEA}$ an inverse output-oriented DEA weight. If $S$ contains
the active monotonicity and concavity rows, the coefficient problem is

\[
\begin{aligned}
\min_{\beta_r,e_r}\quad & \sum_{i\in E_r}w_i e_{ir}\\
\text{s.t.}\quad & B_E\beta_r-y_E=e_r,\\
& e_r\geq0,\\
& S\beta_r\geq0,\\
& \beta_r\ \text{free}.
\end{aligned}
\]

The residual $e_{ir}$ is excess fitted output above observation $i$. Its
non-negativity gives the envelopment property. DEA weights give more influence
to observations nearer the frontier. In a multi-output fit, all outputs share
the selected basis structure, but this program is solved separately for each
output, so the coefficient vectors can differ.

```{r optimization-controls, eval=FALSE}
# These arguments determine the basis and the shape rows of the LP.
fit_lp <- aces(
  production_data,
  x = 1:3,
  y = 4,
  shape = list(mono = TRUE, conc = TRUE),
  max_terms = 50,
  err_red = 0.01,
  kn_penalty = 2
)
```

Because $e_r=B_E\beta_r-y_E$, the error variables can be eliminated. After
dropping a constant from the objective, the exact reduced primal used by the
linear implementation is

\[
\begin{aligned}
\min_{\beta_r}\quad & c^\top\beta_r\\
\text{s.t.}\quad & A\beta_r\geq b,\qquad \beta_r\ \text{free},
\end{aligned}
\]

where $c=B_E^\top w$, $A=[B_E;S]$, and $b=[y_E;0]$. Its dual is

\[
\max_{\lambda\geq0} b^\top\lambda
\quad\text{s.t.}\quad A^\top\lambda=c.
\]

```{r reduced-primal, eval=FALSE}
# Schematic construction; ACES builds these matrices internally.
cvec <- crossprod(B_E, w)
Amat <- rbind(B_E, S)
bvec <- c(y_E, rep(0, nrow(S)))

# min cvec' beta, subject to Amat %*% beta >= bvec; beta is free
```

The package first solves the smaller non-negative dual and recovers the primal
coefficients from solver dual values. It then checks primal feasibility and
strong duality, including the solver sign convention. If that recovery fails,
the reduced primal is solved directly with GLPK. Multiple coefficient vectors
can be optimal, so fitted outputs, feasibility, and the objective value are more
useful reproducibility checks than exact coefficient equality.

```{r compare-fits, eval=FALSE}
tech_a <- get_technology(fit_a, method = "aces")
tech_b <- get_technology(fit_b, method = "aces")

max(abs(tech_a - tech_b))
all.equal(
  get_scores(production_data, 1:3, 4, object = fit_a),
  get_scores(production_data, 1:3, 4, object = fit_b),
  tolerance = 1e-6
)
```

This LP is solved repeatedly during the forward search. Runtime consequently
grows with the number of iterations, inputs, knots, interactions, outputs, and
RF learners. For an exploratory run, the most effective controls are a smaller
`max_terms`, a larger `err_red`, an additive basis, a coarser knot grid, or Quick
ACES.

```{r computational-budget, eval=FALSE}
fit_fast <- aces(
  production_data,
  x = 1:3,
  y = 4,
  quick_aces = TRUE,
  mul_BF = list(max_degree = 1, inter_cost = 0.05),
  max_terms = 30,
  err_red = 0.03
)
```

# Building the production technology

## Multiple outputs and joint feasibility

ACES uses a common basis structure for all outputs but estimates separate
coefficients. The following two-input, two-output example is small enough to
show the complete workflow.

```{r multiple-output-fit, eval=FALSE}
set.seed(314)
multi_data <- translog_X2Y2(N = 60, border = 0.1, noise = FALSE)

multi_fit <- aces(
  data = multi_data,
  x = 1:2,
  y = 3:4,
  mul_BF = list(max_degree = 2, inter_cost = 0.05)
)
```

Estimating each output component separately can produce a vector that looks
plausible component by component but is not jointly attainable. ACES therefore
compares the joint radial score $\phi_i$ with each single-output score
$\phi_i^{(r)}$. A predicted component is retained when

\[
|\phi_i^{(r)}-\phi_i|\leq\psi;
\]

otherwise the observed component is used. A larger (psi) accepts more
predicted components. In a single-output model, the two scores coincide and the
prediction is retained.

`change_technology()` lets the analyst change this refinement rule without
refitting the splines. The supplied matrices must correspond to the rows used by
the fitted model.

```{r refinement, eval=FALSE}
multi_fit_psi <- change_technology(
  object = multi_fit,
  tech_xmat = as.matrix(multi_data[, 1:2]),
  tech_ymat = as.matrix(multi_data[, 3:4]),
  psi = 0.05
)
```

## Reference points and feasible activities

`get_technology()` returns the input and refined-output points that define the
selected technology. It does not enumerate every feasible activity in the
production set.

```{r extract-technology, eval=FALSE}
technology <- get_technology(multi_fit, method = "aces")
head(technology)
```

Under variable returns to scale, the envelopment of reference points
$(x_i,\widetilde y_i)$ is

\[
\widehat T_{VRS}=\left\{(x,y):
x\geq\sum_i\lambda_i x_i,\ 
y\leq\sum_i\lambda_i\widetilde y_i,\ 
\mathbf 1^\top\lambda=1,\lambda\geq0\right\}.
\]

The inequalities give free disposability and the intensity variables give a
convex technology. Under constant returns to scale, the convexity equation is
removed and the set becomes conical. The returns-to-scale assumption is chosen
when scores and targets are computed.

The origin can be added as an explicit reference point with `pto = TRUE`. This
asserts that zero inputs can jointly produce zero outputs and changes the convex
hull; it is a production assumption, not a plotting convenience.

```{r origin, eval=FALSE}
multi_fit_pto <- change_technology(multi_fit, pto = TRUE)
get_technology(multi_fit_pto, method = "aces")
```

# Measuring efficiency

There is no universally best efficiency measure. The measure should reflect the
adjustment that managers can actually contemplate. Output orientation asks how
much outcomes could improve with resources fixed. Input orientation asks how
much resources could fall with outcomes fixed. Non-radial measures allow the
components to move at different rates.

| `measure` | Question answered | Efficient value |
|:--|:--|:--|
| `rad_out` | By what common factor can all outputs expand? | 1 |
| `rad_inp` | By what common factor can all inputs contract? | 1 |
| `ddf` | How far can the DMU move in a chosen direction? | 0 |
| `rsl_out`, `rsl_inp` | What if outputs or inputs adjust separately? | 1 |
| `wam_mip`, `wam_nor`, `wam_ram`, `wam_bam` | How large are normalized input and output slacks? | 0 |
| `rf_aces_rad_out` | What radial score follows directly from RF predictions? | 1 |

## Radial output and input measures

The default output-oriented model maximizes a common expansion factor
$\phi_o$:

\[
\max_{\phi_o,\lambda}\ \phi_o
\quad\text{s.t.}\quad
X\lambda\leq x_o,\quad
Y\lambda\geq\phi_o y_o,
\]

with the selected returns-to-scale constraint. A score of 1 is efficient. A
score of 1.20 means that every output could increase by 20% along this radial
path while inputs remain fixed.

```{r rad-out, eval=FALSE}
radial_output <- get_scores(
  eval_data = production_data,
  x = 1:3,
  y = 4,
  object = fit,
  method = "aces",
  measure = "rad_out",
  returns = "variable"
)
head(radial_output)
```

The input-oriented model instead minimizes $\theta_o$, subject to
$X\lambda\leq\theta_o x_o$ and $Y\lambda\geq y_o$. A value of 0.80 means
that a 20% proportional input contraction is feasible along the chosen path.

```{r rad-inp, eval=FALSE}
radial_input <- get_scores(
  production_data, 1:3, 4,
  object = fit,
  measure = "rad_inp"
)
```

## Directional and non-radial measures

The directional distance function moves from $(x_o,y_o)$ to
$(x_o-\beta_o g_x,y_o+\beta_o g_y)$ and maximizes $\beta_o$. Each row of
`direction` contains input directions first and output directions second. Since
the direction determines the units and the relative management trade-off, it
must be reported with the score.

```{r ddf, eval=FALSE}
direction <- cbind(
  as.matrix(production_data[, 1:3]),
  as.matrix(production_data[, 4])
)

directional <- get_scores(
  production_data, 1:3, 4,
  object = fit,
  measure = "ddf",
  direction = direction
)
```

Russell measures estimate a different expansion factor for each output or a
different contraction factor for each input. They reveal non-radial
inefficiency that one common factor can hide.

```{r russell, eval=FALSE}
russell_output <- get_scores(
  multi_data, 1:2, 3:4,
  object = multi_fit,
  measure = "rsl_out"
)

russell_input <- get_scores(
  multi_data, 1:2, 3:4,
  object = multi_fit,
  measure = "rsl_inp"
)
```

Weighted additive measures work directly with input excesses and output
shortfalls. `wam_mip` scales by the evaluated DMU's values, `wam_nor` by sample
standard deviations, `wam_ram` by sample ranges, and `wam_bam` by distances to
observed bounds. These normalizations answer different comparability questions,
so they need not produce the same ranking.

```{r weighted-additive, eval=FALSE}
wam_scores <- lapply(
  c("wam_mip", "wam_nor", "wam_ram", "wam_bam"),
  function(m) {
    get_scores(
      multi_data, 1:2, 3:4,
      object = multi_fit,
      measure = m
    )
  }
)
```

## Returns to scale and relevant inputs

VRS compares a DMU with convex combinations of reference activities at a
comparable scale. CRS additionally permits proportional scaling of those
activities. The difference between CRS and VRS scores combines technical and
scale effects; it should not be read as a causal effect of becoming larger.

```{r returns-to-scale, eval=FALSE}
score_vrs <- get_scores(
  production_data, 1:3, 4,
  object = fit,
  returns = "variable"
)

score_crs <- get_scores(
  production_data, 1:3, 4,
  object = fit,
  returns = "constant"
)
```

With `relevant = TRUE`, the envelopment problem retains only inputs represented
in the selected basis. Interactions are traced back to their original inputs;
for RF-ACES the union across learners is used. This is a structural post-fit
rule. It does not use a permutation-importance threshold and it does not refit
the frontier.

```{r relevant-inputs, eval=FALSE}
scores_all_inputs <- get_scores(
  production_data, 1:3, 4,
  relevant = FALSE,
  object = fit
)

scores_selected_inputs <- get_scores(
  production_data, 1:3, 4,
  relevant = TRUE,
  object = fit
)
```

## From a score to a target

`get_targets()` solves the same envelopment model and returns the associated
input and output coordinates with suffix `_hat`. A target is conditional on the
chosen method, orientation, returns to scale, direction, and reference sample.
It is a benchmark, not a forecast or an automatic operational commitment.

```{r targets, eval=FALSE}
targets_out <- get_targets(
  eval_data = production_data,
  x = 1:3,
  y = 4,
  object = fit,
  method = "aces",
  measure = "rad_out",
  returns = "variable"
)

head(targets_out)
```

The arguments used for a target should be identical to those used for the
reported score. The simplest way to guarantee that is to keep them in one list.

```{r target-consistency, eval=FALSE}
efficiency_spec <- list(
  eval_data = multi_data,
  x = 1:2,
  y = 3:4,
  object = multi_fit,
  method = "aces_cubic",
  measure = "ddf",
  returns = "variable",
  direction = cbind(multi_data[, 1:2], multi_data[, 3:4])
)

scores_ddf  <- do.call(get_scores, efficiency_spec)
targets_ddf <- do.call(get_targets, efficiency_spec)
```

# Quick ACES

Quick ACES is not a different production model. It is a faster way to search for
an ACES model. It reduces work in three places and then returns the same object
structure used by the standard downstream functions.

First, it screens prepared inputs with Spearman and Kendall rank correlations.
An input is retained when at least one association is sufficiently positive for
at least one output. Rank correlations capture monotone nonlinear relationships
better than Pearson correlation, but screening can still miss an input whose
contribution is mainly local or interactive.

Second, an initial DEA assessment concentrates knot candidates near efficient
observations and their neighboring input values. Those observations are the
most informative about the boundary, so this is more targeted than simply using
a uniformly coarse grid.

Third, the forward search keeps an exponentially weighted history of recent
error reductions. Later candidate evaluations are concentrated on inputs that
have continued to improve the fit. More precisely, at each accepted iteration
ACES records, for every evaluated input, the largest relative loss reduction
among all of its candidate basis-function pairs. The input does not need to
supply the pair that enters the model. This is precisely the information used
by Quick ACES to allocate the next candidate search.

```{r quick-fit, eval=FALSE}
quick_fit <- aces(
  data = production_data,
  x = 1:3,
  y = 4,
  quick_aces = TRUE,
  max_terms = 35,
  err_red = 0.02
)

quick_scores <- get_scores(
  production_data, 1:3, 4,
  object = quick_fit
)
```

## Three views of ACES variable importance

The forward history is available as a training-search importance diagnostic.
`aces_varimp()` adds the best reductions across forward iterations and, by
default, rescales the largest total to 100. This avoids assigning all the credit
to the variable that happened to win each iteration. It should not be read as
test-set importance, statistical significance, or a causal effect.

With `importance = "forward"`, the default, `aces_varimp()` does not refit ACES
or solve any additional optimization problem. The candidate models have already
been evaluated during the forward step of `aces()`, which stores the best
reduction found for each prepared input in every accepted iteration. The
function only retrieves that matrix and either returns it directly or adds its
columns to form the ranking. It is therefore essentially immediate and does not
modify the fitted object. An object created with an older package version must
be fitted again once because it does not contain this history.

```{r aces-importance, eval=FALSE}
# No new model is fitted: this only summarizes information stored in quick_fit
importance <- aces_varimp(quick_fit)

# Exact relative reduction retained for every iteration and prepared input
reduction_history <- aces_varimp(
  quick_fit,
  control = list(type = "iterations")
)
```

Permutation importance asks a different question: how much does the fitted
model deteriorate when the values of one original input are disconnected from
the outputs? The model is held fixed, so no ACES optimization is repeated.
Interactions are rebuilt after each permutation; permuting `capital`, for
example, also changes `capital:labour`. Repeating the permutation reduces the
Monte Carlo variability of the result.

```{r aces-permutation-importance, eval=FALSE}
permutation_importance <- aces_varimp(
  quick_fit,
  importance = "permutation",
  control = list(
    repeats = 20,
    seed = 314
  )
)
```

The approach-specific arguments live inside `control`. This keeps, for example,
`repeats` and `seed` out of forward and LOCO calls. The list is validated against
the selected approach, so a misplaced option raises an error instead of being
ignored. The fitted representation can be changed in the same list.

```{r aces-permutation-raw, eval=FALSE}
permutation_cubic_raw <- aces_varimp(
  quick_fit,
  importance = "permutation",
  normalize = FALSE,
  control = list(
    method = "aces_cubic",
    repeats = 50,
    seed = 2026
  )
)
```

LOCO asks how much fit is lost when one original input is unavailable to the
whole modelling procedure. ACES is refitted without that input and without
every interaction containing it. This is more computationally demanding than
permutation: a model with three original inputs requires three additional ACES
fits. It can also redistribute signal among correlated inputs because the
remaining model is allowed to adapt.

```{r aces-loco-importance, eval=FALSE}
loco_importance <- aces_varimp(
  quick_fit,
  importance = "loco"
)
```

LOCO normally suppresses the progress printed by its repeated fits. Set
`verbose` inside its own control list when those iterations should be visible.

```{r aces-loco-verbose, eval=FALSE}
loco_cubic <- aces_varimp(
  quick_fit,
  importance = "loco",
  control = list(
    method = "aces_cubic",
    verbose = TRUE
  )
)
```

By default, permutation and LOCO compare the configured loss on the training
observations. This is appropriate for describing how the fitted frontier used
the available sample, and training comparisons use the same DEA-based weights
and loss metric as ACES. They can nevertheless be optimistic if the intended
interpretation is predictive relevance. Supply untouched validation data for
that purpose. Permutation still keeps the original fit fixed; LOCO refits on the
original training sample and evaluates both the complete and reduced models on
the supplied validation observations.

```{r aces-validation-importance, eval=FALSE}
permutation_validation <- aces_varimp(
  quick_fit,
  importance = "permutation",
  control = list(
    eval_data = validation_data,
    repeats = 20,
    seed = 314
  )
)

loco_validation <- aces_varimp(
  quick_fit,
  importance = "loco",
  control = list(eval_data = validation_data)
)
```

For permutation and LOCO, `loss_increase` is the signed difference from the
complete-model loss. With `normalize = TRUE`, negative increases are set to zero
and the largest remaining value is scaled to 100. Use `normalize = FALSE` to
inspect the loss change on its original metric scale. Forward importance ranks
prepared inputs, including explicit interaction terms; permutation and LOCO
rank original inputs and include all their associated interactions.

In Quick ACES, a missing history value means that an input was not evaluated in
that iteration because of screening or adaptive search. It contributes zero to
the cumulative ranking, and `iterations_evaluated` reports how often the input
was actually compared. Standard ACES evaluates a broader candidate set and is
therefore preferable when the ranking itself is the main inferential object.

Because Quick ACES evaluates fewer candidates, it can select a different basis
from standard ACES. That is expected. Compare both specifications on validation
data when the difference matters to the substantive conclusion.

```{r compare-quick, eval=FALSE}
fit_full  <- aces(production_data, 1:3, 4, quick_aces = FALSE)
fit_quick <- aces(production_data, 1:3, 4, quick_aces = TRUE)

score_full  <- get_scores(production_data, 1:3, 4, object = fit_full)
score_quick <- get_scores(production_data, 1:3, 4, object = fit_quick)
```

# RF-ACES

## Fit and predict with a randomized ensemble

RF-ACES repeatedly fits ACES to bootstrap samples and random subsets of
candidate inputs. Averaging is useful when individual learners contain signal
but are not perfectly correlated [@breiman2001; @espana2024rf].

`learners` sets the maximum forest size, `bag_size` the bootstrap sample size,
and `max_feats` the number of inputs considered by each learner. `max_feats`
should be an integer between one and the number of inputs; smaller values create
more diverse learners.

```{r rf-fit, eval=FALSE}
set.seed(314)
rf_fit <- rf_aces(
  data = production_data,
  x = 1:3,
  y = 4,
  learners = 100,
  bag_size = 50,
  max_feats = 1,
  early_stopping = list(
    ma_window = 10,
    tolerance = 5
  )
)

rf_fit
```

An observation omitted from a learner's bootstrap sample is out-of-bag (OOB)
for that learner. RF-ACES uses OOB error to monitor the ensemble. `ma_window`
smooths the error history and `tolerance` counts how many non-improving
iterations are allowed before stopping. Setting a tolerance larger than the
requested forest size effectively prevents early stopping within that run.

```{r rf-stopping, eval=FALSE}
rf_fixed_budget <- rf_aces(
  production_data, 1:3, 4,
  learners = 80,
  early_stopping = list(ma_window = 5, tolerance = 100)
)

rf_adaptive <- rf_aces(
  production_data, 1:3, 4,
  learners = 200,
  early_stopping = list(ma_window = 10, tolerance = 5)
)
```

`predict.rf_aces()` averages learner frontier predictions. It estimates maximum
attainable output at new input values, not expected observed output. The new data
need the original input names but do not need observed outputs.

```{r rf-predict, eval=FALSE}
rf_frontier <- predict(
  object = rf_fit,
  newdata = production_data,
  x = 1:3,
  method = "rf_aces"
)

head(rf_frontier)
```

The available RF representations are `"rf_aces"`, `"rf_aces_cubic"`, and
`"rf_aces_quintic"`. Standard ACES currently exposes its fitted technologies
and efficiency tools but does not export an analogous `predict.aces()` method.

## Two ways to obtain an RF efficiency score

The ensemble stores a production technology, so all standard envelopment scores
and targets work with an RF object.

```{r rf-technology, eval=FALSE}
rf_technology <- get_technology(rf_fit, method = "rf_aces")

rf_scores <- get_scores(
  production_data, 1:3, 4,
  object = rf_fit,
  method = "rf_aces",
  measure = "rad_out"
)

rf_targets <- get_targets(
  production_data, 1:3, 4,
  object = rf_fit,
  method = "rf_aces",
  measure = "rad_out"
)
```

There is also a direct RF radial score,

\[
\widehat\phi_o=\min_r
\frac{\widehat f_r(x_o)}{y_{or}},
\]

available as `measure = "rf_aces_rad_out"`. It compares RF frontier predictions
with observed outputs and avoids a new envelopment optimization. It is not
identical to `"rad_out"`, which evaluates the DMU against the aggregated
technology. Choose one definition and use it consistently in a ranking.

```{r rf-direct-score, eval=FALSE}
rf_ratio_score <- get_scores(
  production_data, 1:3, 4,
  object = rf_fit,
  method = "rf_aces",
  measure = "rf_aces_rad_out"
)
```

## Which inputs matter to the forest?

The same three questions used for ACES can be asked of an RF-ACES fit. The
forward measure describes the internal search, permutation measures the
sensitivity of the fitted forest, and LOCO measures what is lost when the whole
forest is rebuilt without an input. They share the `aces_varimp()` interface,
but their computational costs and interpretations remain different.

Every RF-ACES learner stores the best reduction found for each candidate input
at every accepted forward iteration. Within a learner these reductions are
added; the forest score is the average of those totals across all learners.
Inputs not evaluated by a learner contribute zero to that learner, because they
did not have an opportunity to improve its model. No learner is refitted.

```{r rf-forward-importance, eval=FALSE}
rf_forward <- aces_varimp(rf_fit)

rf_forward_history <- aces_varimp(
  rf_fit,
  control = list(type = "iterations")
)
```

The history includes `learner` and `iteration` before the prepared input
columns. In the overall ranking, `learners_evaluated` and
`iterations_evaluated` show how widely each input was actually considered. This
is especially useful when `max_feats` or Quick ACES makes the candidate search
sparse.

For permutation, OOB evaluation is the RF-ACES default. Each observation is
predicted only by learners whose bootstrap sample excluded it, and those
learner predictions are averaged before the loss is calculated. The fitted
forest stays fixed. Permuting an original input also rebuilds every interaction
that contains it.

```{r rf-permutation-oob, eval=FALSE}
rf_permutation_oob <- aces_varimp(
  rf_fit,
  importance = "permutation",
  control = list(
    method = "rf_aces",
    eval_data = "oob",
    repeats = 20,
    seed = 314
  )
)
```

The representation is selected with the same `method` option used elsewhere in
the package. RF-ACES offers `"rf_aces"`, `"rf_aces_cubic"`, and
`"rf_aces_quintic"`. Setting `normalize = FALSE` retains the signed change in
the fitted loss on its original scale.

```{r rf-permutation-method, eval=FALSE}
rf_permutation_cubic_raw <- aces_varimp(
  rf_fit,
  importance = "permutation",
  normalize = FALSE,
  control = list(
    method = "rf_aces_cubic",
    eval_data = "oob",
    repeats = 50,
    seed = 2026
  )
)
```

Untouched data can replace OOB evaluation. The data frame or matrix must keep
the model variables in their original columns. All learners then predict every
validation observation, and the validation observations receive equal weight.

```{r rf-permutation-validation, eval=FALSE}
rf_permutation_validation <- aces_varimp(
  rf_fit,
  importance = "permutation",
  control = list(
    method = "rf_aces",
    eval_data = validation_data,
    repeats = 20,
    seed = 314
  )
)
```

LOCO refits a complete forest after removing each original input and all its
interactions. With OOB evaluation, the full and reduced losses for each input
use only observations that have an OOB prediction in both forests. It is the
most complete of the three measures, but also the most expensive: three inputs
require three additional RF-ACES fits. Since those fits are randomized, use
enough learners to limit Monte Carlo noise and call `set.seed()` immediately
before LOCO when the ranking must be reproducible.

```{r rf-loco-oob, eval=FALSE}
set.seed(314)
rf_loco_oob <- aces_varimp(
  rf_fit,
  importance = "loco",
  control = list(
    method = "rf_aces",
    eval_data = "oob"
  )
)
```

The reduced forests are always fitted on the original training sample. An
external data set changes only where the complete and reduced forests are
compared.

```{r rf-loco-validation, eval=FALSE}
rf_loco_validation <- aces_varimp(
  rf_fit,
  importance = "loco",
  control = list(
    method = "rf_aces",
    eval_data = validation_data,
    verbose = TRUE
  )
)
```

Permutation and LOCO measure predictive dependence, not causality or
statistical significance. Correlated inputs can divide, mask, or replace one
another's contribution. These rankings are also separate from
`relevant = TRUE`, which looks at the inputs represented structurally in fitted
basis functions.

# Intervals from RF-ACES

## What the learner spread tells us

`rf_aces_intervals()` evaluates the requested quantity separately for every
learner and takes empirical quantiles. For output intervals, the learner values
are frontier predictions. For technology-based score intervals, the efficiency
problem is resolved against every learner-specific technology. The latter is
important: uncertainty is propagated through both frontier estimation and the
efficiency optimization.

```{r rf-ensemble-intervals, eval=FALSE}
ensemble_intervals <- rf_aces_intervals(
  object = rf_fit,
  newdata = production_data,
  x = 1:3,
  y = 4,
  level = 0.95,
  type = "both",
  method = "rf_aces",
  measure = "rad_out",
  returns = "variable",
  calibration = "none"
)

head(ensemble_intervals$output)
head(ensemble_intervals$score)
```

Here, `level = 0.95` means that the limits are the 2.5th and 97.5th percentiles
of learner values. It does **not** by itself mean that the interval has 95%
coverage for a future observation or for the true frontier. Learners are
correlated, their spread is not the entire sampling distribution, and systematic
frontier bias or observational noise may not appear in that spread. This is the
precise sense in which the interval is *uncalibrated*.

Even so, an ensemble interval is useful. A narrow interval says that the fitted
answer is stable across the bootstrap samples and random feature choices used by
the forest. A wide interval warns that the benchmark is sensitive to those
modelling perturbations.

## Calibrating intervals for observed outputs

With `calibration = "oob"`, RF-ACES uses absolute OOB residuals and a
finite-sample conformal quantile to enlarge the output interval. Under
exchangeability, this targets approximate marginal coverage for a future
**observed output**. `min_oob` is the minimum number of OOB predictions required
before a training observation contributes a calibration residual.

```{r rf-calibrated-output, eval=FALSE}
prediction_interval <- rf_aces_intervals(
  object = rf_fit,
  newdata = production_data,
  x = 1:3,
  y = 4,
  level = 0.95,
  type = "output",
  method = "rf_aces",
  calibration = "oob",
  min_oob = 5
)

prediction_interval$interval_type
prediction_interval$calibration
head(prediction_interval$output)
```

This calibration target should not be overstated. It is an approximate marginal
prediction interval, not a conditional interval for every input vector. With
several outputs, coverage is marginal for each component rather than simultaneous
for the whole vector. Most importantly, it concerns future observed output, not
the latent production frontier.

True unit-level efficiencies are not observed, so ordinary OOB residuals cannot
calibrate an efficiency-score interval in the same way. Score intervals remain
ensemble stability intervals even when output calibration is requested.

```{r rf-score-interval, eval=FALSE}
score_interval <- rf_aces_intervals(
  object = rf_fit,
  newdata = production_data,
  x = 1:3,
  y = 4,
  level = 0.90,
  type = "score",
  method = "rf_aces",
  measure = "rad_out",
  relevant = FALSE,
  returns = "variable"
)

head(score_interval$score)
```

# Ranking units across RF-ACES learners

`ranking_aces()` calculates one efficiency score per learner and then averages
those scores for each unit. It does not average rank positions. Preserving the
score scale uses the magnitude of the efficiency differences and makes the
final order easier to interpret.

For training units, the default OOB ranking uses only learners whose bootstrap
sample did not contain the unit. The unit therefore influenced neither the
learner fit nor its reference technology. This is preferable to evaluating the
training observations against every learner, which would mix honest OOB scores
with optimistic in-bag scores.

```{r rf-ranking-oob, eval=FALSE}
training_ranking <- ranking_aces(
  object = rf_fit,
  eval_data = "oob",
  method = "rf_aces",
  measure = "rad_out",
  returns = "variable"
)

head(training_ranking)
```

The result reports the average score, its standard deviation across eligible
learners, and `learners_used`. OOB counts can differ between units because every
bootstrap sample omits a different set of observations. `score_sd` is a
descriptive stability measure, not a standard error or confidence interval.

External units were not used by any learner, so all learners can contribute.
The data and its input-output indexes are kept together in one named list.

```{r rf-ranking-external, eval=FALSE}
external_ranking <- ranking_aces(
  object = rf_fit,
  eval_data = list(
    data = validation_data,
    x = 1:3,
    y = 4
  ),
  method = "rf_aces",
  measure = "rad_out"
)

head(external_ranking)
```

This learner-average ranking is generally different from calling `get_scores()`
once on the aggregated RF technology. The former solves one efficiency problem
per learner and then averages the solutions; the latter aggregates the reference
technology first and solves only once. Both are meaningful, but they answer
different questions.

The order automatically follows the selected measure. For `rad_out`,
`rsl_out`, `ddf`, the weighted additive measures, and `rf_aces_rad_out`, smaller
scores are better. For `rad_inp` and `rsl_inp`, larger scores are better. Ranks
are formed before rounding, so the `digits` argument cannot introduce artificial
ties.

```{r rf-ranking-input-oriented, eval=FALSE}
input_ranking <- ranking_aces(
  object = rf_fit,
  eval_data = "oob",
  method = "rf_aces_cubic",
  measure = "rad_inp",
  relevant = FALSE,
  digits = 4
)
```

The same function can rank external units with a standard ACES fit. In that
case there is one technology rather than a learner distribution, so
`learners_used` equals one and `score_sd` is unavailable.

```{r aces-ranking-external, eval=FALSE}
aces_ranking <- ranking_aces(
  object = fit,
  eval_data = list(
    data = validation_data,
    x = 1:3,
    y = 4
  ),
  method = "aces",
  measure = "rad_out"
)
```

# Validation and tuning

## Reuse the same assessment splits

`data_partition()` stores the analysis and assessment indexes so that every
parameter configuration is compared on the same observations. It supports
`"kfold"`, `"repeated_kfold"`, `"holdout"`, `"bootstrap"`, `"loo"`, and
`"group_kfold"`.

```{r partition, eval=FALSE}
partition <- data_partition(
  data = production_data,
  x = 1:3,
  y = 4,
  validation = "kfold",
  resampling_args = list(folds = 5),
  delta = 1.05,
  seed = 42
)
```

For each split, the assessment error is weighted by inverse output-oriented DEA
scores computed from the analysis sample. `delta` removes assessment observations
whose score lies too far beyond the analysis technology. This keeps validation
focused on frontier prediction instead of allowing many interior observations
to dominate an average regression error.

Random folds are inappropriate when they leak information across time, sites, or
related units. In those settings, use groups or construct assessment periods that
match the intended deployment.

```{r partition-alternatives, eval=FALSE}
repeated <- data_partition(
  production_data, 1:3, 4,
  validation = "repeated_kfold",
  resampling_args = list(folds = 5, repeats = 3),
  delta = 1.10,
  seed = 42
)

holdout <- data_partition(
  production_data, 1:3, 4,
  validation = "holdout",
  resampling_args = list(train_prop = 0.8),
  seed = 42
)
```

## Tune ACES without losing the full validation history

`tune_model_aces()` evaluates the Cartesian product of the supplied values.
Nested lists such as `shape` and `mul_BF` are grids as well. The first validation
metric chooses the default best configuration, but every requested metric is
stored. Available metrics are `mae`, `mape`, `mse`, `msle`, `rmse`, `rmsle`,
`nrmse_mean`, and `nrmse_range`.

```{r tune-aces, eval=FALSE}
tuned_aces <- tune_model_aces(
  partition = partition,
  mul_BF = list(
    max_degree = c(1, 2),
    inter_cost = c(0.01, 0.05)
  ),
  shape = list(
    mono = TRUE,
    conc = c(TRUE, FALSE)
  ),
  max_terms = c(30, 50),
  err_red = c(0.01, 0.03),
  validation_metric = c("rmse", "mae"),
  returns = "variable",
  method = "aces",
  verbose = TRUE
)

tuned_aces
tuned_aces$best_params
tuned_aces$results
```

The tuning function refits the selected configuration to the complete data and
stores it in `final_model`. `select_best()` can select another configuration by
any metric that was already computed. This is helpful when, for example, RMSE
and MAE imply different operational priorities.

```{r select-best, eval=FALSE}
final_fit <- tuned_aces$final_model

best_by_mae <- select_best(tuned_aces, metric = "mae")
best_by_mae$best_params
best_by_mae$best_scores
best_by_mae$config_id
```

`tune_model_rf_aces()` uses the same partition and adds the ensemble controls.
The grid can grow very quickly, so start with a small range of substantively
plausible values and expand around competitive regions.

```{r tune-rf, eval=FALSE}
tuned_rf <- tune_model_rf_aces(
  partition = partition,
  learners = c(50, 100),
  bag_size = c(40, 50),
  max_feats = c(1, 2),
  early_stopping = list(
    ma_window = c(5, 10),
    tolerance = c(3, 5)
  ),
  max_terms = 30,
  err_red = c(0.01, 0.03),
  validation_metric = "rmse",
  method = "rf_aces",
  verbose = TRUE
)

tuned_rf$best_params
```

# Simulation tools

The package includes several data-generating processes for checking frontier
recovery, variable selection, noise sensitivity, and multi-output behavior. When
a generator returns both `y` and `yT`, only `y` is observed by the estimator;
`yT` is retained for simulation evaluation.

## Cobb-Douglas designs

`cobb_douglas_XnY1()` provides one output with 1, 3, 6, 9, 12, or 15 inputs.
`cobb_douglas_XnY1_noise()` adds statistical noise and can generate a
heteroskedastic three-input design. `cobb_douglas_X3Y3()` supplies three inputs
and three outputs under increasing, constant, or decreasing returns.

```{r simulations-cd, eval=FALSE}
sim_cd <- cobb_douglas_XnY1(N = 200, nX = 6)

sim_noise <- cobb_douglas_XnY1_noise(
  N = 200,
  nX = 3,
  p = 0.5,
  heteroskedasticity = TRUE
)

sim_3x3 <- cobb_douglas_X3Y3(
  N = 200,
  border = 0.2,
  noise = 0.5,
  returns = "DRS"
)
```

`cobb_douglas_XnHnY1()` adds `nH` irrelevant inputs to `nX` productive inputs.
The correlation structures are supplied as named matrices. This design is
particularly useful for comparing Quick ACES screening, structural relevance,
and RF permutation importance.

```{r simulation-irrelevant, eval=FALSE}
rho_x <- matrix(
  0, nrow = 3, ncol = 3,
  dimnames = list(paste0("x", 1:3), paste0("x", 1:3))
)
rho_x["x1", "x2"] <- 0.3

rho_h <- matrix(
  0, nrow = 3, ncol = 3,
  dimnames = list(paste0("x", 1:3), paste0("h", 1:3))
)
rho_h["x1", "h1"] <- 0.3

sim_irrelevant <- cobb_douglas_XnHnY1(
  N = 300,
  nX = 3,
  nH = 3,
  exp_x = rep(1 / 3, 3),
  rho_x = rho_x,
  rho_h = rho_h
)
```

## Additive and multi-output designs

`add_scenario_XnY1()` implements six additive scenarios, labelled A-F.
Comparing them reveals how the method responds to changes in dimension and
functional shape.

```{r simulation-additive, eval=FALSE}
sim_add_A <- add_scenario_XnY1(N = 200, scenario = "A")
sim_add_F <- add_scenario_XnY1(N = 200, scenario = "F")
```

`translog_X2Y2()` is a compact two-input, two-output design for studying joint
refinement and non-radial efficiency. `cet_cd_X2Y2()` combines a Cobb-Douglas
input function with a constant-elasticity-of-transformation output function and
contains producer groups operating in different scale regions.

```{r simulations-multi, eval=FALSE}
sim_translog <- translog_X2Y2(
  N = 250,
  border = 0.1,
  noise = FALSE
)

sim_cet <- cet_cd_X2Y2(
  N = 400,
  border = 0.1,
  noise = FALSE
)
```

`mult_scenario_XnY1()` is intended to provide the multiplicative counterpart of
the A-F scenarios. In package version 0.2.0, it references an undefined internal
object while constructing `y` and does not return a data set. Its interface can
be inspected, but it should not be used in a reproducible analysis until the
implementation is corrected.

```{r simulation-multiplicative-status, eval=FALSE}
args(mult_scenario_XnY1)

# Currently not operational:
# sim_mult_F <- mult_scenario_XnY1(N = 200, scenario = "F")
```

# Interpreting and reporting the results

An efficiency result is relative to the observed reference group and to every
modelling choice that defines the benchmark. A score of one means that the DMU
lies on the estimated frontier under that specification. It does not prove
absolute best practice, nor does it establish that the unit has no room to
improve outside the observed production system.

A practical sensitivity analysis changes one substantive assumption at a time:
returns to scale, method, input set, orientation, refinement threshold, or ACES
variant. Conclusions that reverse under small defensible changes should be
reported as fragile.

```{r sensitivity, eval=FALSE}
sensitivity_results <- list(
  aces_vrs = get_scores(
    production_data, 1:3, 4,
    object = fit, measure = "rad_out", returns = "variable"
  ),
  aces_crs = get_scores(
    production_data, 1:3, 4,
    object = fit, measure = "rad_out", returns = "constant"
  ),
  aces_vrs_relevant = get_scores(
    production_data, 1:3, 4,
    object = fit, relevant = TRUE, measure = "rad_out"
  )
)
```

Targets also need an operational review. The mathematical projection may ignore
budgets, service obligations, quality thresholds, indivisibilities, or local
environmental constraints. It is best used to quantify a performance gap and
identify relevant peers before it is translated into a management commitment.

```{r management-table, eval=FALSE}
management_table <- cbind(
  production_data,
  score = get_scores(production_data, 1:3, 4, object = fit)[[1]],
  get_targets(production_data, 1:3, 4, object = fit)
)

head(management_table)
```

New input combinations outside the training support are extrapolations. Shape
restrictions make that extrapolation coherent, but not empirically supported.
At minimum, flag observations outside each training range; for an important
decision, inspect multivariate support as well.

```{r support-check, eval=FALSE}
train_x <- production_data[, 1:3]
new_x   <- new_producers[, 1:3]

outside_support <- vapply(seq_along(train_x), function(j) {
  new_x[[j]] < min(train_x[[j]]) | new_x[[j]] > max(train_x[[j]])
}, logical(nrow(new_x)))

new_producers$outside_marginal_support <- rowSums(outside_support) > 0
```

A reproducible study should report the DMU definition and sampling period;
input and output definitions and units; transformations; ACES variant and shape
restrictions; interaction and knot controls; refinement threshold; fitted
method; returns to scale; efficiency measure, orientation, direction, or
weights; validation design; and the sensitivity or interval analysis used.

```{r reporting-record, eval=FALSE}
analysis_specification <- list(
  estimator = "ACES",
  inputs = names(production_data)[1:3],
  outputs = names(production_data)[4],
  method = "aces",
  shape = list(mono = TRUE, conc = TRUE),
  max_degree = 1,
  psi = 0.05,
  returns = "variable",
  measure = "rad_out",
  relevant = FALSE,
  validation = "5-fold frontier-weighted CV"
)
```

# A complete workflow

The following template keeps model selection, final fitting, efficiency, and
targets as distinct steps. A small exploratory fit should precede a larger grid
so that data and solver problems are found early.

```{r complete-workflow, eval=FALSE}
library(aces)

# Comparable DMUs and variables
x <- match(c("labour", "capital", "materials"), names(data))
y <- match(c("service", "quality"), names(data))

# Reusable frontier-aware assessment splits
partition <- data_partition(
  data, x, y,
  validation = "repeated_kfold",
  resampling_args = list(folds = 5, repeats = 3),
  delta = 1.05,
  seed = 2026
)

# A deliberately small initial grid
tuning <- tune_model_aces(
  partition,
  mul_BF = list(
    max_degree = c(1, 2),
    inter_cost = c(0.03, 0.05)
  ),
  max_terms = c(30, 50),
  err_red = c(0.01, 0.03),
  validation_metric = c("rmse", "mae"),
  method = "aces"
)

# The selected specification is already refitted to all observations
final_fit <- tuning$final_model

# Freeze the benchmark definition before producing results
final_scores <- get_scores(
  data, x, y,
  object = final_fit,
  method = "aces",
  measure = "rad_out",
  returns = "variable"
)

final_targets <- get_targets(
  data, x, y,
  object = final_fit,
  method = "aces",
  measure = "rad_out",
  returns = "variable"
)

results <- cbind(data, final_scores, final_targets)
```

# Advanced and developmental functions

`aces()` and `rf_aces()` are the user-facing wrappers. They validate the input,
prepare and scale the data, run the algorithm, construct technologies, and
return objects that work with the rest of the package. The exported
`aces_algorithm()` and `rf_aces_algorithm()` functions are lower-level engines
for package development. Calling them directly requires the already prepared
internal structures, so ordinary analyses should use the wrappers.

```{r low-level-functions, eval=FALSE}
# Recommended user interface
fit    <- aces(production_data, 1:3, 4)
rf_fit <- rf_aces(production_data, 1:3, 4)

# Useful when extending the package
args(aces_algorithm)
args(rf_aces_algorithm)
```

# References

The production-function foundation of ACES is developed in @espana2024. The
multi-output technology and refinement procedure are presented in @espana2025,
and the randomized estimator in @espana2024rf. General production-frontier and
DEA foundations include @fare1994 and @banker1984.
